\Large{All in the family:\\German twin finishing times in\\the 2016 women's Olympic marathon}

### 2017-02-25 paper/chance_memo.tex
edits prior to submitting revision

\noindent
\begin{tabular}{ll}
  \textbf{To:} &  Scott Evans, editor, \emph{CHANCE}\\
  \textbf{From:} & David Cottrell and Michael C.\ Herron, Dartmouth College\\
  \textbf{Date:} & \today \\
  \textbf{Re:} &  Revisions to women's marathon submission
\end{tabular}

@@

  The $t$-value for the quadratic term is -1.88 and the corresponding
  $p$-value is 0.062, which is not particularly compelling.  We have
  accordingly dropped the quadratic terms in the scatter plots on p.\
  4 and, implicitly, from the Studentized residual plots on p.\ 6.  In
  terms of the latter suggestion about, Figure 1b contains a plot of
  half marathon split times versus actual finishing times.
  %% Note: xr labels are
  %% breaking; looks like some sort of
  %% label clash with another package.
  %% Hardcoding here is poor style.

@@

>>>>

\noindent
\begin{tabular}{ll}
  \textbf{To:} &  Scott Evans, editor, \emph{CHANCE}\\
  \textbf{From:} & David Cottrell and Michael C.\ Herron\\
  \textbf{Date:} & \today \\
  \textbf{Re:} &  Revisions to women's marathon submission
\end{tabular}

@@

  The $t$-value for the quadratic term is -1.88 and the corresponding
  $p$-value is 0.062, which is not particularly compelling.  We have
  accordingly dropped the quadratic terms in the scatter plots on p.\
  4 and, implicitly, from the Studentized residual plots on p.\ 6.
  Regarding the latter suggestion concerned half marathon splits and
  observed finishing times, Figure 1b contains such a plot.
  %% Note: xr labels are
  %% breaking; looks like some sort of
  %% label clash with another package.
  %% Hardcoding here is poor style.

@@

  The blindfolding question is addressed in Figures 5b and 6b.  In
  short, a consecutive Hahner twin finish assuming independence (that
  is, blindfoldedness) would have been very unlikely.

### 2017-02-25 paper/paper.tex
edits prior to submitting revision

The media quickly picked up the Hanher story as an image of the
beaming twins, finishing hand-in-hand, captured a public
audience. While many believed the twin's near-simultaneous finish was
a reflection of Olympic spirit, not everyone agreed with this rosy
interpretation. The twins' happy facial expressions at the finish were
portrayed as a bit contrived---smiling like ``Honigkuchenpferde,''
cookies in the shape of a horse, was the description offered by one
editorialist---and the sports director of the German Athletics
Federation, Thomas Kurschilgen, stirred up controversy when he
suggested that the Hahners' photo-finish was no coincidence.
Kurschilgen averred that the twins slowed down so as to finish
simultaneously and create a spectacle which would ``generate media
attention.''  Kurschilgen justified his charge with the fact that the
twins ran the Rio marathon over 18 minutes slower their personal best
times prior to the Olympics.  Kurschilgen's accusations were denied by
the Hahner twins, perhaps not surprisingly, who claimed that their
joint finish was simply a fortuitous coincidence.

@@

For each participant who started the women's Olympic marathon, we know
several things: personal best marathon time prior to the 2016 Olympic
games; age; split times from the Rio marathon course at 5 kilometers,
10 kilometers, and so forth; and, overall finishing time. We cannot
directly observe the effort that an individual put into the race, and
we do not know why some runners have DNF results.  Some runners may
have injured themselves on the course and accordingly dropped out, and
others may have dropped out, uninjured, in anticipation of an
unsatisfactory result. Of the 156 marathon starters, 133 completed the
race and 23 DNFed at various locations throughout the course. The
overall DNF rate was thus $\frac{23}{156} \approx 0.15$, and the
relatively small sample size at our disposal yields a relatively wide
95\% confidence for this rate, namely, $\left(0.098, 0.22\right)$.

@@

Our research design is twofold.  First, we present visualizations that
describe various features of the 2016 women's Olympic marathon, and an
important element in our visualizations is the difference between
runners' Rio times and their prior personal best marathon times.  Our
visualizations suggest that the Hahner twin's pace in the marathon was
slow, albeit not excessively so, but that their simultaneous finish
was quite unusual given the twins' differences in abilities (and
similarly for the Kim twins).  We then turn to a regression-based
simulation of the marathon, and our simulations reinforce what we
observed in prior visualizations, namely, that the Hahner twins did
not run appreciably slowly yet finished suspiciously close to each
other.  We return in our conclusion to Kurschilgen's claims about the
Hahners and offer thoughts about their validity.

@@

One way that we might assess whether a women marathoner's Rio finish
was unusual---or, say, whether two finishes were jointly unusual---is
by comparing a runner's observed Olympic finishing time on August 14,
2016, with a measure of her underlying marathon talent.  For the
latter we use prior personal best marathon times.  This is a natural
measure of marathon talent, we believe, but it is not entirely free of
complications.  For example, personal best times are potentially
confounded by the marathon courses on which they were set; some
courses, like the Berlin marathon, are known for relatively fast
times.  In addition, personal best times may be confounded by
conditions, like weather, that varied across races.  Finally, personal
best times may not capture Olympic raceday idiosyncrasies that could
affect individual runners.  For example, a runner could have woken up
with a minor cold on the morning of August 14, 2016.  With these
concerns in mind, we use a runner's half marathon split time from the
Olympic marathon as a secondary measure of athletic skill in
marathoning.  Our two measures of marathon talent are featured in the
visualizations below.

Figure \ref{fig:scatter} contains two plots, both of which describe
how finishing times from the Rio women's marathon varied as a function
of athletes' personal best times and half marathon split times.  The
points in the plots are colored by twin/triplet status, and both plots
contain %second-order polynomial regression lines.
dashed lines representing least squares regression fits.  The ticks
along the horizontal axes in both plots indicate times, either
personal bests or half marathon splits, of runners who earned DNF
results.\label{scatterplots}

@@

With respect to our second measure of marathon skill, Figure
\ref{fig:45degreeplot_half} shows that there was nothing abnormal
about the Hahner twins' overall finishing times, conditional on their
half marathon splits. As one might expect, a runner's time halfway
through the course is a fairly strong predictor of her finishing time.
Here we see that the relationship between the Hahner twin's half
marathon split times and their finishing times is similar to that of
other Rio runners.  The two points representing the Hahners are in the
middle of the distribution of points and therefore imply that the
twins do not seem to have slowed down in any unusual way.  This is
consistent with the previous Figure \ref{fig:45degreeplot} and
inconsistent with accusations made against the Hahner twins, at least
the part of the accusation that focused on their overall pace.

@@

The personal best times of the Hahner twins were 115 seconds apart and
their official finishing times were separated by one second.  Is such
a 115 to one compression typical among pairs of runners?  Moreover,
are there other pairs of Olympic marathoners who had a difference
between personal best times of 115 seconds apart and, if so, how close
were their finishing times?

@@

We can generalize this result by looking at all pairs of runners in
the marathon.  For all 8,778 pairs of 133 finishers, Figure
\ref{fig:diffdiffscatter} plots differences in finishing times against
differences in personal best times, and pairs of twins/triplets are
identified by the same color scheme we used earlier.  Consider first
the Hahner twins.  The two German women are effectively located on the
horizontal axis because their difference in finishing times is one
second.  However, there are many points above the Hahner's black dot,
and this shows that, conditional on an approximate 115 second
difference in personal best times, most marathoners did not have close
finishing times like the Hahners.  Some pairs of runners with around
115 second personal best differences had finishing time differences of
1,000 seconds, i.e., in excess of 15 minutes.  The points in Figure
\ref{fig:diffdiffscatter} are not independent, but they provide a
rough sense of the dispersion in finishing time differences between
runners that one might expect conditional on differences in personal
best times.

@@

Figure \ref{fig:secondsbehind} describes each Olympic runner's status
at various split times on the marathon course.  Each dot in the
figure---color as before---depicts a recorded split and the number of
seconds each runner was behind the race leader at the time.  There are
more dots at earlier splits due to subsequent DNFs.

@@

The Estonian triplet DNF prior to the half marathon split is evident
in Figure \ref{fig:secondsbehind} , which also shows that Lisa Hahner
was ahead of her sister through 15 kilometers.  The figure contains
two red dots representing the North Korean Kim twins, but this is not
visually apparent because the Kim twins had identical split times
during the entire marathon.

@@

To estimate the conditional distribution of each Rio final result, we
assume that each runner's marathon time \Yi\ is distributed normally
with a mean that is a function of the runner's marathon
skill - which we assume can be determined by a linear combination of her personal best marathon time $X_i$ and her age $Z_i$. Hence:
$$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i}  + \beta_2\,Z_{i} + e_i, \sigma\right),$$
%%
where $N\left(\cdot,\cdot\right)$ denotes a normal
distribution. We estimate $\beta_0$,
$\beta_1$, $\beta_2$, and $\sigma$ using ordinary least squares on
finishing Rio marathoners.  We exclude the Hahner/Kim twins and Luik
triplets from the sample so that that our estimates are not affected
by the twin/triplet finishes which, theoretically, could reflect
runner coordination.  For a simulated marathon, we draw a runner's
time from an estimated distribution and condition on the runner's
personal best marathon time and age.  Once a race is simulated for all
runners, twins and triplets included, we record both the time between
Anna and Lisa Hahner's simulated finishes and the difference in their
simulated ranks.  We then simulate a new race---drawing a new set of
finishing times---and record the same quantities. The steps of the
simulation are as follows.
\begin{enumerate}
\item Ignoring twins and triplets, estimate with least squares a
  linear model that predicts a runner's finishing time $Y_i$ based on
  her pre-Olympic personal best time $X_i$ and her age $Z_i$.
\item Extract the resulting coefficient vector, estimated covariance
  matrix, and estimated regression variance from this model.
\item For each simulated rate, draw intercept and slope estimates
  $\tilde{\beta}_0$, $\tilde{\beta}_1$, and
  $\tilde{\beta}_2$, respectively, from a multivariate normal
  distribution with mean equal to the previously estimated coefficient
  vector and covariance equal to the previously estimated covariance
  matrix.
\item For each runner, draw an error $\tilde{e}$ from a normal
  distribution with mean zero and a standard deviation equal to the
  standard deviation of the original regression model's residuals.
\item Predict each runner's final result by combining the randomly
  generated beta coefficients and individual error terms,
  $\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1\,X_i +  \tilde{\beta}_2\,Z_i + \tilde{e}$.
\item Eliminate each runner from the simulated race with a probability
  equal to the observed fraction of marathoners who did not finish the
  marathon.
% \item Calculate the difference in time and difference in ranking
%   between Anna Hahner and Lisa Hahner, assuming both finished the
%   race.
\item Repeat above steps 10,000 times.
\end{enumerate}

As these steps illustrate, our simulation repeatedly draws random
coefficient vectors, and this captures variability in what we know
about the relationship between runner talent and age and runner finish
times.  In addition, the simulation draws random disturbances for each
runner, conditional on our original estimate of regression variance;
these disturbances capture variability in runner finishing times,
notwithstanding underlying marathon talent as proxied for by prior
best finishing time.  Importantly, the disturbances that we draw are
independent across runners.  Consequently, for each simulated race the
finishing order among runners will vary.

From our simulations we generate intervals which describe the extent
of the variability in marathon finishing times. For example, in 95\%
of the simulations in which Anna Hahner completed the marathon, her
finishing time was between 8,369 seconds and 9,999 seconds. This is
consistent with Anna's observed finishing time in Rio, which was 9,932
seconds. Nonetheless, her finishing time was on the slower end of this interval.   
In fact, our simulations estimate that, given Anna's personal best marathon time
and her age, she would be expected to finish the Rio marathon in 9,181 seconds, which is slightly more
than 12 minutes faster than her actual time.  This is similar to her sister's result. Lisa Hahner's
corresponding 95\% interval is 8,507 to 10,143 with a mean of 9,319.  Her observed
finishing time of 9,933 seconds.

The bottom line here is that our simulated marathon finishes are
consistent with the Hahner twins' finishing times in that their finishes were inside
the bounds of traditional 95\% confidence intervals. And note that the
regression model underlying the simulations was estimated without the
Hahner twins (and same for the Kim twins and Luik triplets). According
to our simulation, then, both Hahner twins did not run appreciably
slowly, conditional on personal best times prior to the Rio Olympics
and age.

@@

The histograms in Figure \ref{fig:simdiff} raise questions about the
credibility of Anna and Lisa Hahner's story and in particular suggest
that a simultaneous finish in the Rio marathon would be very rare if
Anna and Lisa had run independently. For example, in fewer than 300 of
10,000 simulated races did Anna and Lisa Hahner finish within 30
seconds of one another, and in fewer than 600 did the Hahner twins
finish within a minute of each other.  The histogram area associated
with this result is depicted in red in Figure
\ref{fig:simulatedfinishtimes}.  Moreover, the Hahner twins finished
in consecutive rank in fewer than 200 of 10,000 simulated races; the
red zone in Figure \ref{fig:simulatedranks} presents this visually.
The close finish that was observed in Rio, where Anna and Lisa crossed
the finish line one after the other, would have been highly unlikely
if the two German twins had raced independently of each other.

@@

We have offered two perspectives on the marathon, one based on
visualizations and simple calculations and a second that draws on a
simulation. Both perspectives have the same implications, and they are
as follows. In a global sense, the Hahner twins did not slow down
appreciably during the Rio marathon. Their times were not fast, but
they were within reason for runners of the Hahners' abilities.
Locally, though, we find that the Hahner's finish was in fact
contrived. Their finish---one second between the two women---was an
extremely low probability event. Compared to their differences in
talent, the Hahner twins difference in finishing times was unusually
compressed. This is evidence that their finishing had elements of
intentionality---perhaps at the last minute but intentionality
nonetheless.

Our goal is not to speak to the question of whether the Hahner twins
should or should not have enjoyed a somewhat contrived moment of
Olympic glory. Neither was in contention for a podium finish in Rio,
and compared to the doping allegations that presently surround
endurance sports in general, what the Hahners did seems relatively
tame. Still, it might have behooved them to be a bit more open about
their end-of-race tactics, and to this end we have shown here how a
simple data analysis can shed light on claims about racing results.
Should the 2020 Summer Olympics feature twins and triplets again, we
look forward to a comparative analysis using the techniques
illustrated here.

>>>>

The media quickly picked up the Hanher story as an image of the
beaming twins, finishing hand-in-hand, captured a public
audience. While many believed the twin's near-simultaneous finish was
a reflection of Olympic spirit, not everyone agreed with this rosy
interpretation. The twins' happy facial expressions at the finish were
portrayed as a bit contrived---smiling like ``Honigkuchenpferde,''
cookies in the shape of a horse, was the description offered by one
editorialist---and the sports director of the German Athletics
Federation, Thomas Kurschilgen, stirred up controversy when he
suggested that the Hahners' photo-finish was no coincidence.
Kurschilgen averred that the twins slowed down so as to finish
simultaneously and create a spectacle, and he justified his charge by
noting that the Hahners ran the Rio marathon over 18 minutes slower their
personal best times prior to the Olympics.  Kurschilgen's accusations
were denied by the Hahner twins, perhaps not surprisingly, who claimed
that their joint finish was simply a fortuitous coincidence.

@@

For each participant who started the women's Olympic marathon, we know
several things: personal best marathon time prior to the 2016 Olympic
games; age; split times from the Rio marathon course at 5 kilometers,
10 kilometers, and so forth; and, overall finishing time. We cannot
directly observe the effort that an individual put into the race, and
we do not know why some runners have DNF results.  Some runners may
have injured themselves on the course and accordingly dropped out, and
others may have stopped running, uninjured, in anticipation of an
unsatisfactory result. Of 156 marathon starters, 133 completed the
race and 23 DNFed at various locations throughout the course. The
overall DNF rate was thus $\frac{23}{156} \approx 0.15$, and the
relatively small sample size at our disposal yields a relatively wide
95\% confidence for this rate, namely, $\left(0.098, 0.22\right)$.

@@

Our research design is twofold.  First, we present visualizations that
describe various features of the 2016 women's Olympic marathon, and
among other things our visualizations feature difference between
runners' Rio times and their prior personal best marathon times.  The
visualizations suggest that the Hahner twin's pace in the marathon was
slow, albeit not excessively so, but that their simultaneous finish
was quite unusual given the twins' differences in abilities (and
similarly for the Kim twins).  We then turn to a regression-based
simulation of the marathon, and our simulations reinforce what we
observed in prior visualizations, namely, that the Hahner twins did
not run appreciably slowly yet finished suspiciously close to each
other.  We return in our conclusion to Kurschilgen's claims about the
Hahners and offer thoughts about their validity.

@@

One way that we might assess whether a women marathoner's Rio finish
was unusual---or, say, whether two finishes were jointly unusual---is
by comparing a runner's observed Olympic finishing time on August 14,
2016, with a measure of her underlying marathon talent.  For the
latter we use prior personal best marathon times.  This is a natural
measure of marathon talent, we believe, but it is not entirely free of
complications.  For example, personal best times are potentially
confounded by the marathon courses on which they were set; some
courses, like the Berlin marathon, are known for relatively fast
times.  In addition, personal best times may be confounded by
conditions, like weather, that varied across races.  Finally, personal
best times may not capture Olympic raceday idiosyncrasies that could
affect individual runners.  For example, a runner could have woken up
with a minor cold on the morning of August 14, 2016.  With these
concerns in mind, we use a runner's half marathon split time from the
Olympic marathon as a secondary measure of athletic skill in
marathoning.

Figure \ref{fig:scatter} contains two plots describing how finishing
times from the Rio women's marathon varied as a function of athletes'
personal best times and half marathon split times.  The points in the
plots are colored by twin/triplet status, and both plots
contain %second-order polynomial regression lines.
dashed lines representing least squares regression fits.  The ticks
along the horizontal axes in both plots indicate times, either
personal bests or half marathon splits, of runners who earned DNF
results.\label{scatterplots}

@@

With respect to our second measure of marathon skill, Figure
\ref{fig:45degreeplot_half} shows that there was nothing abnormal
about the Hahner twins' overall finishing times, conditional on their
half marathon splits. As one might expect, a runner's time halfway
through the course is a fairly strong predictor of her finishing time.
Here we see that the relationship between the Hahner twin's half
marathon split times and their finishing times is similar to that of
other Rio runners.  The two points representing the Hahners are in the
middle of the distribution of points and therefore imply that the
twins do not seem to have slowed down dramatically after their reached
the halfway point of the Rio marathon.  This is consistent with the
aforementioned Figure \ref{fig:45degreeplot} and inconsistent with
accusations made against the Hahner twins, at least the part of the
accusation that focused on their overall pace.

@@

The personal best times of the Hahner twins were 115 seconds apart and
their official finishing times were separated by one second.  Is such
a 115 to one compression typical among pairs of runners?  Relatedly,
are there other pairs of Olympic marathoners who had a difference
between personal best times of 115 seconds apart and, if so, how close
were their finishing times?

@@

We can generalize this result by looking at all pairs of runners in
the marathon.  For all 8,778 pairs of 133 finishers, Figure
\ref{fig:diffdiffscatter} plots differences in finishing times against
differences in personal best times, and pairs of twins/triplets are
identified by the same color scheme we used earlier (recall that only
two of the Luik triplets finished the Rio marathon).  Consider first
the Hahner twins.  The two German women are effectively located on the
horizontal axis because their difference in finishing times is one
second.  However, there are many points above the Hahner's black dot,
and this shows that, conditional on an approximate 115 second
difference in personal best times, most marathoner pairs did not have
close finishing times like the Hahners.  Some pairs of runners with
around 115 second personal best differences had finishing time
differences of 1,000 seconds, i.e., in excess of 15 minutes.  The
points in Figure \ref{fig:diffdiffscatter} are not independent, but
they provide a rough sense of the dispersion in finishing time
differences between runners that one might expect conditional on
differences in personal best times.

@@

Figure \ref{fig:secondsbehind} describes each Olympic runner's status
at various split times on the marathon course.  Each dot in the
figure---colored as before---depicts a recorded split and the number of
seconds each runner was behind the race leader at the time.  There are
more dots at earlier splits due to subsequent DNFs.

@@

The Estonian triplet DNF prior to the half marathon split is evident
in Figure \ref{fig:secondsbehind} , which also shows that Lisa Hahner
was ahead of her sister through 15 kilometers.  The figure contains
two red dots representing the North Korean Kim twins, but this is not
visually apparent because the Kim twins had identical split times
during the entire marathon.  This explains why we used the term
``lockstep'' to refer to the Kim twin's marathon pace.

@@

To estimate the conditional distribution of each Rio final result, we
assume that each runner's marathon time \Yi\ is distributed normally
with a mean that is a function of the runner's marathon skill, which
we assume is a linear combination of her personal best marathon time
$X_i$ and her age $Z_i$:
$$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i}  + \beta_2\,Z_{i} + e_i, \sigma\right),$$
%%
where $N\left(\cdot,\cdot\right)$ denotes a normal
distribution. We estimate $\beta_0$,
$\beta_1$, $\beta_2$, and $\sigma$ using ordinary least squares on
finishing Rio marathoners.  We exclude the Hahner/Kim twins and Luik
triplets from the sample so that that our estimates are not affected
by the twin/triplet finishes which, theoretically, could reflect
runner coordination.  For a simulated marathon, we draw a runner's
time from an estimated distribution and condition on the runner's
personal best marathon time and age.  Once a race is simulated for all
runners, twins and triplets included, we record both the time between
Anna and Lisa Hahner's simulated finishes and the difference in their
simulated ranks.  We then simulate a new race---drawing a new set of
finishing times---and record the same quantities. The steps of the
simulation are as follows.
\begin{enumerate}
\item Ignoring twins and triplets, estimate with least squares a
  linear model that predicts a runner's finishing time $Y_i$ based on
  her pre-Olympic personal best time $X_i$ and her age $Z_i$.
\item Extract the resulting coefficient vector, estimated covariance
  matrix, and estimated regression variance from this model.
\item For each simulated race, draw intercept and slope estimates
  $\tilde{\beta}_0$, $\tilde{\beta}_1$, and
  $\tilde{\beta}_2$, respectively, from a multivariate normal
  distribution with mean equal to the previously estimated coefficient
  vector and covariance equal to the previously estimated covariance
  matrix.
\item For each runner, draw an error $\tilde{e}$ from a normal
  distribution with mean zero and a standard deviation equal to the
  standard deviation of the original regression model's residuals.
  This step requires 156 draws from a normal distribution, one draw
  per marathoner.
\item Predict each runner's final result by combining the randomly
  generated beta coefficients and individual error terms,
  $\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1\,X_i +  \tilde{\beta}_2\,Z_i + \tilde{e}$.
\item Eliminate each runner from the simulated race with a probability
  equal to the observed fraction of marathoners who did not finish the
  marathon.
% \item Calculate the difference in time and difference in ranking
%   between Anna Hahner and Lisa Hahner, assuming both finished the
%   race.
\item Repeat above steps 10,000 times.
\end{enumerate}

As these steps illustrate, our simulation repeatedly draws random
coefficient vectors, and this captures uncertainty in what we know
about the relationship between runner talent and age and runner finish
times.  In addition, for each simulated race our simulation draws
random disturbances for each runner, conditional on the original
estimate of regression variance; these disturbances capture
variability in runner finishing times, notwithstanding age and
underlying marathon talent.  Importantly, the disturbances that we
draw are independent across runners.  Consequently, for each simulated
race the finishing order among runners will vary.

From our simulations we generate intervals which describe the extent
of the variability in marathon finishing times. For example, in 95\%
of the simulations in which Anna Hahner completed the marathon, her
finishing time was between 8,369 seconds and 9,999 seconds. This is
consistent with Anna's observed finishing time in Rio, which was 9,932
seconds. Nonetheless, Anna's finishing time was on the slower end of
this interval.  In fact, our simulations estimate that, given Anna's
personal best marathon time and her age, she would be expected to
finish the Rio marathon in 9,181 seconds, which is slightly more than
12 minutes faster than her actual time.  This is similar to her
sister's result. Lisa Hahner's corresponding 95\% interval is 8,507 to
10,143 with a mean of 9,319.  Her actual finishing time was 9,933
seconds.

The bottom line here is that our simulated marathon finishes are
consistent with the Hahner twins' finishing times in that their
finishes were inside the bounds of traditional 95\% intervals. And
note that the regression model underlying the simulations was
estimated without the Hahner twins (and same for the Kim twins and
Luik triplets). According to our simulation, then, both Hahner twins
did not run appreciably slowly, conditional on personal best times
prior to the Rio Olympics and age.

@@

The histograms in Figure \ref{fig:simdiff} raise questions about the
credibility of Anna and Lisa Hahner's story and in particular suggest
that a simultaneous finish in the Rio marathon would be very rare if
Anna and Lisa had run independently. For example, in fewer than 300 of
10,000 simulated races did Anna and Lisa Hahner finish within 30
seconds of one another, and in fewer than 600 did the Hahner twins
finish within a minute of each other.  The histogram area associated
with this latter result is depicted in red in Figure
\ref{fig:simulatedfinishtimes}.  Moreover, the Hahner twins finished
in consecutive rank in fewer than 200 of 10,000 simulated races; the
red zone in Figure \ref{fig:simulatedranks} presents this visually.
The close finish that was observed in Rio, where Anna and Lisa crossed
the finish line one after the other, would have been highly unlikely
if the two German twins had raced independently of each other.

@@

We have offered two perspectives on the marathon, one based on
visualizations and simple calculations and a second that draws on a
simulation. Both perspectives have the same implications, and they are
as follows. In a global sense, the Hahner twins did not slow down
appreciably during the Rio marathon. Their times were not fast, but
they were within reason for runners of the Hahners' abilities and age.
Locally, though, we find that the Hahner's finish was in fact
contrived. Their finish---consecutive, with one second between the two
women---was a rather low probability event. Compared to their
differences in talent, that is, the Hahner twins difference in
finishing times was unusually compressed. This is evidence that their
finishing had elements of intentionality---perhaps at the last minute
but intentionality nonetheless.

Our goal is not to speak to the question of whether the Hahner twins
should or should not have enjoyed a somewhat contrived moment of
Olympic glory. Neither was in contention for a podium finish in Rio,
and compared to the doping allegations that presently surround
endurance sports in general, what the Hahners appear to have done
seems relatively tame. Still, it might have behooved them to have been
a bit more open about their end-of-race tactics, and to this end we
have shown here how a simple data analysis can shed light on claims
about racing results.  Should the 2020 Summer Olympics again feature
marathoning twins and triplets, we look forward to a comparative
analysis using the techniques illustrated here.
